---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  fig.width = 6,
  fig.asp = 0.618,
  out.width = "70%",
  fig.align = "center",
  dpi = 300
)
```

# Notes: Loading the Crossref dump into Google BigQuery with R

In April 2020, Crossref made a data file of its index freely available via Academic Torrents

<https://doi.org/10.13003/83B2GP>

Here, I describe how I downloaded and parsed a subset from the dump, comprising journal articles published since 2008, and loaded it into Google BiqQuery for large-scale analysis.

## Aim

Make use of the Crossref dump for publisher and journal-specific analysis focusing on

- Publication volume
- License information
- Links to full texts for text mining

for journal articles published since 2008.

Complements the API route to access Crossref metadata.

## Methods

### Download Dump

Download from [Academic Torrents](http://academictorrents.com/details/0c6c3fbfdc13f0169b561d29354ea8b188eb9d63) via the command line:

```shell
$ wget http://academictorrents.com/download/0c6c3fbfdc13f0169b561d29354ea8b188eb9d63.torrent
$ ctorrent 0c6c3fbfdc13f0169b561d29354ea8b188eb9d63.torrent
```

Duration: 45 minutes. 

The dump itself consists of 37501 json-formatted files à 3.000 Crossref records.

### Data transformation

Considerations:

- With rcrossref, a well-tested and widely used parser for Crossref metadata already exists
- Only a subset of metadata is needed for the analysis
- Already Split files makes it possible to process dump files independently from each other

Pipeline:

- [rcrossref_parser.R](rcrossref_parser.R) copy of internal parser functions from rcrossref
- [cr_dump_parser.R](cr_dump_parser.R) use of rcrossref parser in specific pipeline for processing the dump files
- [read_and_write.R](https://github.com/njahn82/cr_dump/blob/master/read_and_write.R#L32) apply parser on each dump file in multiple sessions using future and future.apply package

Machine: [Amazon EC2 c5.4xlarge](https://aws.amazon.com/ec2/instance-types/c5/).


Lessons learned:

- Checking for relevant files before applying the parser significantly speed-ups the process!
- Tracking the parsing process is critical (files parsed, and cpu usage). A great tool is [htop](https://hisham.hm/htop/).
- Process should be stopped and re-started after some CPUs have finished their task

### Upload transformed data to Google BiqQuery

Pre-processing:

- Google BQ does not allow `.` in field names. I had to changed it using the bash script [bash_sed_bq.sh](bash_sed_bq.sh) (Absolute file paths are not a good practise)

Upload to Google Cloud Storage in parallel:

```shell
$ gsutil -m cp -r data_parsed gs://cr_dump
```

Load into BQ

```shell
bq load --ignore_unknown_values 
        --source_format=NEWLINE_DELIMITED_JSON 
        api-project-764811344545:cr_dump_march_20.cr_08_20 
        gs://cr_dump/data_parsed/*.json schema_cr.json
```
 
Upload and loading was the quickest, around 15 minutes upload time, and 75s to populate the BQ table.

## Use cases

```{r}
library(tidyverse)
# connect to google bq
library(DBI)
library(bigrquery)
con <- dbConnect(
  bigrquery::bigquery(),
  project = "api-project-764811344545",
  dataset = "cr_dump_march_20"
)
bq_auth()
```

### Journal article volume by publisher (Top 10)

```{sql, connection=con}
SELECT member, publisher, COUNT(DISTINCT(doi)) as n
FROM `api-project-764811344545.cr_dump_march_20.cr_08_20` 
GROUP BY member, publisher
ORDER BY n DESC
LIMIT 10
```


### Licence penetration

#### Articles with license information

```{sql, connection=con, output.var="license_all"}
SELECT COUNT(DISTINCT(DOI)) as n
FROM `api-project-764811344545.cr_dump_march_20.cr_08_20` 
WHERE ARRAY_LENGTH(license) > 0
```

`r license_all$n` articles published since 2008 have licence information.

#### Creative Commons license URLs by publisher (Top 10)

```{sql, connection= con}
SELECT publisher, COUNT(DISTINCT(DOI)) as n
FROM `api-project-764811344545.cr_dump_march_20.cr_08_20`, unnest(license) as license
WHERE REGEXP_CONTAINS(license.URL, 'creativecommons')
GROUP BY publisher
ORDER BY n DESC
LIMIT 10
```

#### Open Archive Elsevier

```{sql, connection=con, output.var="els_delay"}
SELECT license.delay_in_days, COUNT(DISTINCT(DOI)) as n
FROM `api-project-764811344545.cr_dump_march_20.cr_08_20`, unnest(license) as license
WHERE REGEXP_CONTAINS(license.URL, 'elsevier.com/open-access/userlicense/1.0/')
GROUP BY license.delay_in_days
ORDER BY delay_in_days
```

```{r}
ggplot(els_delay, aes(delay_in_days, n)) +
  geom_density(stat = "identity") 
```

### Text mining

#### How many publisher provide full-text links intended for text mining?

```{sql, connection=con}
SELECT COUNT(DISTINCT(publisher)) as n
FROM `api-project-764811344545.cr_dump_march_20.cr_08_20`, unnest(link) as link
WHERE link.intended_application = 'text-mining' and link.content_version = 'vor'
```

#### What file fomats are supported?

```{sql, connection=con}
SELECT link.content_type, COUNT(DISTINCT(doi)) as n
FROM `api-project-764811344545.cr_dump_march_20.cr_08_20`, unnest(link) as link
WHERE link.intended_application = 'text-mining' and link.content_version = 'vor'
GROUP BY link.content_type
ORDER BY n DESC
LIMIT 10
```

### Limits


- It needs to be validated whether all records were correctly parsed. At least, there's a slight difference between our Unpaywall dump and the imported Crossref dump.

```{sql connection=con}
SELECT COUNT(DISTINCT(upw.doi)) as no_overlap
FROM   `api-project-764811344545.oadoi_full.mongo_upwFeb20_13_20` as upw
WHERE genre = 'journal-article' AND NOT EXISTS (SELECT 1 
                   FROM `api-project-764811344545.cr_dump_march_20.cr_08_20` as cr
                   WHERE  LOWER(cr.doi) = upw.doi)
```

- Journal disambigation needed, maybe using `issn-l`
- Generally, not all crossref metadata fiels were imported


